---
title: "R workshop"
output: html_document
author: Caroline Graham
date: 1/21/2019
---
#Creating objects in R

```{r arithmetic}
3 + 5
12 / 7
5*5

#assign value to object
weight_kg <- 55

2.2 * weight_kg

```


```{r}
library (tidyverse)

surveys <- read_csv("https://ndownloader.figshare.com/files/2292169")

library(here) #this package makes working directores and file paths easy

surveys

str(surveys)

dim(surveys)
nrow(surveys)
ncol(surveys)

summary(surveys)

```

###Indexing and subsetting data frames

```{r setup, warning=FALSE}
#First define the row coordinate, and then the column. Also write row and then column

surveys[1,1]

surveys[1,6]

#Defining only which elements we want will return a data frame

surveys[1]

surveys[1:3, 7]

```

##Data manipulation

Key functions for data manipulation:

* select(): subsetting columns
* filter(): subsets of rows based on conditions
* mutate(): create new columns, based on information from other columns
* group_by(): creates groups based on categorical data in a column
* summarize(): creates summary stats on grouped data
* arrange(): sort results
* count(): gives a count of discrete variables

```{r}
select(surveys, plot_id, species_id, weight)

select(surveys, -record_id) #negative subsetting

filter(surveys, year == 1995)

```
## Pipes

%>% command - shift - m

```{r}

surveys %>% 
  filter(weight >5) %>% 
  select(species_id, sex, weight)

surveys_sml <- surveys %>% 
  filter(weight >5) %>% 
  select(species_id, sex, weight)

surveys_challenge1 <- surveys %>% 
  filter(year == 1995) %>%
  select(year, sex, weight)

```

```{r}
surveys %>% 
  mutate(weight_kg = weight/1000, weight_kg2 = weight_kg *2)

surveys %>% 
  mutate(mean_weight = mean(weight)) 


# this last step didn't work because there were NAs in the weight column

surveys <- surveys %>% 
   drop_na(weight) %>% 
   mutate(mean_weight = mean(weight)) 

mean(surveys$weight) 

challenge2 <- surveys %>% 
  drop_na(hindfoot_length) %>%
  mutate(hindfoot_half = hindfoot_length/2) %>% 
  filter(hindfoot_half < 30) %>%
  select(species_id, hindfoot_half)

```

```{r}
surveys %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE))

surveys %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>% 
  arrange(min_weight)

```

```{r}
surveys %>% 
  count(sex)

#code above is the same as
surveys %>%  
  group_by(sex) %>%  
  summarize(count = n())

surveys %>% 
  count(plot_type)

surveys %>% 
  group_by(species_id) %>% 
  summarize(mean_length= mean(hindfoot_length, na.rm = TRUE),  min_length = min(hindfoot_length, na.rm = TRUE), max_length = max(hindfoot_length, na.rm = TRUE), n())

surveys %>% 
  group_by(species_id, year) %>%  
  summarize(max_weight = max(weight, na.rm = TRUE)) %>% 
  arrange (desc(max_weight))

max_weights <- surveys %>% 
  drop_na(weight) %>%  
  group_by(year) %>% 
  filter (weight == max(weight)) %>% 
  select(year, genus, species, weight) %>% 
  arrange(year)

  
```

##Export our data

```{r}

write_csv(max_weights, here("write_data", "max_weights.csv"))

```

# Day 2

Note that shortcut for <- is option + "-"

## Tidy data in spreadsheets

The functions we use to tidy data are:

* tidyr::spread
* tidyr::gather

### spread()

spread() takes three principal arguments:
1. the data
2. the _key_ column variable will become the new column names
3. the value column variable which will fill the new column variables

We are going to use the surveys dataset


```{r}
install.packages('tinytex') # for converting to PDFs

surveys <- read_csv(here("read_data", "surveys.csv"))

#create a wide data format of surveys using spread

surveys_gw <- surveys %>% 
  drop_na(weight) %>% 
  group_by(species_id) %>% 
  summarize(mean_weight = mean(weight))

str(surveys_gw)
  
# now we want to spread the data set out into wide format

wide_surveys_gw <- surveys_gw %>% 
  spread(key = species_id, value = mean_weight)


```

### gather()

gather() takes four arguments:

1. data
2. key
3. value
4. names of columns we use to fill the key variable (or drop)

```{r}

# now going back to long format from wide

long_surveys <- wide_surveys_gw %>% 
  gather(key = species_id, value = mean_weight)

```

## Sending tidy data

### Change log

* Update your change log with changes to raw data, or other significant changes or additions to project

### Data dictionary

* Create a data dictionary to define our variables

```{r}
tidy_gsi <- read_csv(here::here("read_data", "tidy_gsi.csv"))

```

# Analyzing Data

## Importing from the Hakai Data Portal

Switched to data_wrangling script to import data into our read_data folder

We have chla data, fish data, and sockeye stock ID data. Let's start to analyze these data sets.

```{r}
fish <- read_csv(here::here("read_data", "fish.csv"))
chla <- read_csv(here::here("read_data", "chla.csv"))
tidy_gsi <- read_csv(here::here("read_data", "tidy_gsi.csv"))

```

```{r}
fish %>% 
  count(species)

fish_d09 <- fish %>%
  filter(site_id == "D09") %>% 
  select(hakai_id, jsp_survey_id, seine_id, date, species, site_id, fork_length, weight) %>% 
  mutate(k = (10 ^ 5 * weight) / fork_length ^ 3) %>% 
  drop_na(k)
```

## Factors

```{r}
str(fish_d09)

class(fish_d09$species)

# coerce a column to be a factor do this:

fish_d09$species <- factor(fish_d09$species)

class(fish_d09$species)

levels(fish_d09$species)

```
If you have factors that are numbers, don't try to do math with those factors. Under the hood R will be treating your factor levels as a different number

##Dates

Note that this did not work!!!

read_csv() treats ISO date standards (yyy-mm-dd) as a DATE object
read.csv() treats them as characters which is not ideal

Lubridate is a package for dealing with dates
 
```{r, eval = FALSE}
library(lubridate)

fish_d09$date <- as_date(fish_d09$date)

class(fish_d09$date)

# extract date components
fish_d09 <- fish_d09 %>% 
  dplyr::mutate(year = year(date),
        month = month(date),
        week = week(date),
        yday = yday(date))



```

Lubridate you can do math with dates

* periods
* intervals
* durations

# Joining Data

Data we have:

* chla
* tidy_gsi
* fish_d09

```{r}
left_join(fish_d09, tidy_gsi, by = 'hakai_id')

right_join(fish_d09, tidy_gsi)

so_gsi <- inner_join(fish_d09, tidy_gsi, by = 'hakai_id')

anti_join(fish_d09, tidy_gsi)

```

## ggplot2

To build a ggplot:

ggplot(data = DATA, mapping = aes(mappings))
    + GEOM_FUNCTION()


```{r, eval = FALSE}
ggplot(data = surveys, mapping = aes(species_id, weight)) + geom_point()

ggplot(so_gsi, aes()) +
  geom_point(aes(x= yday, y= fork_length)) +
  geom_smooth(aes(x= yday, y= fork_length), method = lm)
```

Cookbook for R
How to adjust legends, make facets, adjust axes

http://cookbook-r.com/Graphs

Geom list:

https://ggplot2.tidyverse.org/reference/

```{r}
ggplot(so_gsi, aes()) +
  geom_boxplot(aes(x = month, y = weight))

```

